---
title: "Math 530-630 Midterm Exam"
subtitle: "Fall 2016"
author: "Alison Hill & Joel Adams"
date: "11/9/2016"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(include = FALSE, eval = FALSE)
```

#Free Answer Questions

Grading of the free-answer problems will resemble that of the homeworks:

  * 2 points for the initial answer being "in-good-faith"
  * 3 points for the quality of the final answer / discussion.
    * 3 (Exceptional): answer is thorough, concise, and clearly demonstrates ability to analyze and interpret statistics as well as theoretical understanding of statistical concepts
    * 2 (Adequate): answer addresses the question with moderate inaccuracies in analysis and/or interpretation, or offers a correct but incomplete answer 
    * 1 (Inadequate): answer attempts to address question with substantial inaccuracies in analysis and/or interpretation
    * 0 (Insufficient): answer does not attempt to address question or answer is insufficient to grade


##Probability

1. The probability density function of a random variable $X$ is given by:
$$f(x) = \begin{cases} 
  cx, & 0 < x < 4 \\
  0, & \textrm{otherwise}
  \end{cases}$$
  
    a. find $c$
    b. find the cumulative distribution function $F(x)$.
    c. Compute $P(1 <  X < 3)$

```{r}
Continuous, PDF -> CDF

2.5.6 from 2015 textbook.

a) c = 1/8
b) F(x) = 0,       x<0
          x^2/16,  0 <= x <= 4
          1,       x > 4
c) P(1 < x < 3) = 1/2
```


## Sampling

*Hint: This problem reference ideas and methods from Chapter 4*

2. A friend claims that she has drawn a random sample of size 30 from the exponential distribution with $\lambda = \frac{1}{10}$. The mean of her sample is 12.

    a. What is the expected value of a sample mean?
    b. Run a simulation by drawing 1000 random samples, each of size 30 from Exp(1/10), and then compute the mean. What proportion of the sample means are as large or larger than 12?
    c. Is a mean of 12 unusual for a sample of size 30 from Exp(1/10)?

```{r}
This is C&H Chapter 4, problem 12.  p.93
```


## Bootstrapping

*Hint: This problem reference ideas and methods from Chapter 5*

*Hint: If you want to consistently get the same results while you work, consider using set.seed()*

3. Consider a population that has a normal distribution with mean $\mu = 52$, standard deviation $\sigma = 8$.

    a. The sampling distribution of $\bar{X}$ for samples of size 500 will have what distribution, mean, and standard error?

    b. Use R to draw a random sample of size 500 from this population. What is the mean, standard deviation of this sample?

    c. Compute the bootstrap distribution for your sample and note the bootstrap mean and standard error.

    d. Compare the bootstrap distribution to the theoretical sampling distribution by creating a table like Table 5.2 in Chihara and Hesterberg (p. 104).


```{r}
This is C&H Chapter 5, problem 7, MODIFIED!.  p.130
```

##Hypothesis Testing, Confidence Interval

*Hint: These problems reference ideas and methods from Chapters 7, 8*

4. A researcher believes that in recent years women have been getting taller. She knows that 10 years ago the average height of young adult women living in her city was 63 inches. The standard deviation is unknown. She ranomly samples eight young adult women currently residing in her city and measures their heights. The following data are obtained:
>> Height (in inches): 64, 66, 68, 60, 62, 65, 66, 63
  
    a. What is the alternative hypothesis? (In evaluating this experiment, assume a non-directional hypothesis is appropriate because there are insufficient theoretical and empirical bases to warrant a directional hypothesis.)
    b. What is the null hypothesis?
    c. What kind of test are you going to perform?
    d. Calculate the appropriate obtained statistic.
    e. What is the number of degrees of freedom that is associated with this statistic? And (using $\alpha = 0.01$), what is/are the critical value(s)?
    f. Using $\alpha = 0.01$, what is the conclusion? (i.e. do we retain or reject the null hypothesis, and why?)
    g. Calculate the 99\%-confidence interval around the obtained mean.
    
```{r}
Problem 2 from Alisons Psy310 Final Exam Practice Set
a) In recent years the height of women has been changing. Therefore the sample with \bar{X_{obt}} = 64.25 is a random sample from a population where \mu \neq 63

b) It is reasonable to consider the sample with \bar{X_{obt}} = 64.25 is a random sample from a population where \mu = 63

c) single sample t-test on means

d) S = 2.550, t_{obt} = 1.39

e) df = n-1 = 7. t_{crit} = \pm 3.499

f) since | t_{obt} | < | t_{crit} | , we retain the null hypothesis.

g) (61.1, 67.4)

```

5. Suppose you have the following information concerning the two independent groups below, and
you want to test if the means of the two groups differ.

    a. Is the alternative hypothesis directional or nondirectional?
    b. What is the alternative hypothesis?
    c. What is the null hypothesis?
    d. What kind of test are you going to perform?
    e. Using $\alpha$ = 0.05, what is/are the critical value(s)? Degrees of freedom?
    f. Calculate the appropriate test statistic.
    g. Using $\alpha = 0.05$, what is the conclusion?

| Group 1               | Group 2               |
| --------------------- |-----------------------|
| Mean 1 = 15.6         | Mean 2 = 13.2         |
| Standard dev 1 = 1.4  | Standard dev 2 = 2.7  |
| $n_{1}$ = 10          | $n_{2}$ = 10          |


# Multiple-Choice Questions

* A star ( \* ) next to the problem signifies that that problem is extra credit (there are 7).
* Unstarred problems are required
* Starred problems are worth 1 points
* Unstarred problems are worth 2 point.

1. Suppose you have 10 numbers and have computed the mean to be 8.0. You then discover that the last number in the data was entered incorrectly. It was entered as 8.0 when it should have been 4.0. If you replace the incorrect value (8.0) with the correct one (4.0), and recompute the mean, you will obtain a new mean of:

    a. It is impossible to determine
    b. 6.6
    c. 8.6
    d. 7.6
    
```{r}
Answer: d
```

    
2. You have a set of data that have a mean of 50 and a standard deviation of 12. You wish them to have a mean of 65 and a standard deviation of 10, while retaining the shape of the distribution. What values of *a* and *b* in the linear transformation formula *Y = aX + b* will produce a new set of data with the desired mean and standard deviation?

    a. *a* = 0.833, *b* = 23.3
    b. *a* = 1.83, *b* = -23.3
    c. *a* = 1.67, *b* = 23.3
    d. *a* = 0.833, *b* = 46.7
    e. *a* = 1.2, *b* = 22.3

```{r}
Answer: a
```
    
3. Jane had a z-score of 1.75 on her statistics midterm. If the class mean is 65.0, and the class standard deviation is 12.0, what was Jane's raw score?

    a. It is impossible to determine
    b. 66
    c. 86
    d. 76

```{r}
Answer: c
```


4. IQ scores have a distribution that is approximately normal in shape, with a mean of 100 and a standard deviation of 15 in the general population. Assuming the normal distribution is a good approximation, what proportion of the general population has IQ scores between 79.0 and 109.0?

    a. 0.645
    b. 0.745
    c. 0.545
    d. 0.709
    e. None of the above answers are correct

```{r}
Answer: a
```

```{r}
pnorm(109, 100, 15) - pnorm(79, 100, 15)
```

5. \* You have 10 numbers with a sample mean of 9.0 and a sample variance of 11.0. You discover that the last number in the list was recorded as 8.0 when it should have been recorded as 12.0. If you correct your error and correctly recompute the sample variance, what value will you obtain?

    a. 11.71
    b. 11.0
    c. 10.54
    d. 13.01
    e. None of the above answers are correct
    
```{r}
Answer: a
```

6. Incorrectly rejecting a statistical null hypothesis that is true is to commit:

    a. a Type I error
    b. a Type II error
    c. a Beta error
    d. a Gamma error

```{r}
Answer: a
```


7. If $\alpha$ = 0.01, then the probability of a correct acceptance of a true statistical null hypothesis is:

    a. 0.0001
    b. $\beta$
    c. 0.99
    d. 0
    e. It cannot be determined from the information provided

```{r}
Answer: c
```

8. Given the following probability distribution for the random variable *X*, the variance of *X* is:

    a. 3.5563
    b. 3.7524
    c. 2.95
    d. 1.575
    e. 1.8475

| *x* | $P_X(x)$ | 
|:---:|:--------:|
|   1 |     0.1  |
|  2  |     0.15 |
|    3|      0.2 |
|    4|      0.2 |
|    5|     0.35 |



```{r}
Answer: e
```

9. Given the following probability distribution for the random variable *X*, the expected value of *X* is:

    a. 2.48
    b. 3.68
    c. 3.58
    d. 3.48

| *x* | $P_X(x)$ | 
|:---:|:--------:|
|   1 |     0.1  |
|  2  |     0.2  |
|    3|      0.2 |
|    4|      0.12|
|    5|     0.38 |



```{r}
Answer: d
```

10. The sampling distribution of the sample mean based on *N* iid observations

    a. converges asymptotically to a normal distribution in shape under the conditions of the Central Limit Theorem
    b. has a variance of $\sigma^2/N$ for any population distribution
    c. is always exactly normal, for any sample size, when the population distribution is normal 
    d. all of the above answers are correct

```{r}
Answer: d
```

11. One can increase power by:

    a. increasing *N*
    b. decreasing $\alpha$
    c. increasing the effect size
    d. both a and c

```{r}
Answer: d
```

12. Which of the following may be *false*?

    a. $\alpha + \beta = 1$
    b. power + $\beta$ = 1
    c. 1 - power =  $\beta$
    d. 1 - $\beta$ =  power

```{r}
Answer: a
```

13. \* If one draws all possible samples for various values of *N* from the same population of raw scores, as *N* increases:

    a. The standard error of the mean increases
    b. The standard error of the mean stays the same
    c. The standard error of the mean decreases
    d. The standard error of the mean cannot be calculated

```{r}
Answer: c
```

14. If one draws all possible samples for various values of *N* from the same population of raw scores, as *N* increases:

    a. The mean of the sampling distribution of the mean increases
    b. The mean of the sampling distribution of the mean stays the same
    c. The mean of the sampling distribution of the mean decreases
    d. None of the above

```{r}
Answer: b
```

15. In cases where *N* > 1, the relationship between the raw score population standard deviation and the standard error is:

    a. The standard error is greater than the population standard deviation
    b. The standard error is less than the population standard deviation
    c. The standard error equals the population standard deviation
    d. The standard error is the population standard deviation

```{r}
Answer: b
```

16. \* The 95% confidence interval for a mean of 13.0, standard deviation of 1.6 and *N* = 21 is

    a. 12.01 to 13.99
    b. 12.27 to 13.73
    c. 11.05 to 15.05
    d. 12.95 to 14.95

```{r}
Answer: b
```

```{r}
t_crit <- qt(.975, 20)
13 - t_crit*(1.6/sqrt(21))
13 + t_crit*(1.6/sqrt(21))
```

17. Suppose we wish to test the null hypothesis that the population mean is equal to 50. If we do not know the population standard deviation and thus estimate it using the sample standard deviation, which sampling distribution would be most appropriate for the test?

    a. *z*
    b. *t*
    c. $\chi^2$
    d. *p*

```{r}
Answer: b
```

18. The variance can be thought of as:

    a. half the range
    b. the sum of squared deviations from the mean
    c. the average deviation
    d. the average squared deviation from the mean

```{r}
Answer: d
```

19. What would happen to the mean of a distribution of scores if the number 10 is added to each score?

    a. It would stay the same
    b. It increases by 10
    c. It will become 10 times as large
    d. It will increase, but the amount depends on the shape of the distribution

```{r}
Answer: b
```

20. What would happen to the standard deviation of a distribution of scores if the number 10 is added to each score?

    a. It would stay the same
    b. It increases by 10
    c. It will become 10 times as large
    d. It will increase, but the amount depends on the shape of the distribution

```{r}
Answer: a
```

21.  \* What would happen to the mean of a distribution of scores if each score is multiplied by 2?

    a. It would stay the same
    b. It increases by 2
    c. It will become twice as large
    d. It will increase, but the amount depends on the shape of the distribution

```{r}
Answer: c
```

22. \* What would happen to the variance of a distribution of scores if each score is multiplied by 2?

    a. It would stay the same
    b. It will become twice as large
    c. It will become four times as large
    d. It will increase, but the amount depends on the shape of the distribution

```{r}
Answer: c
```

23. A researcher has data for two variables, *x* and *y*. First, she converts both variables to z-scores with a mean of 0 and standard deviation of 1, and calls them $z_x$ and $z_y$. Next, she takes the mean of both z-scores and calls that new variable $ave_z$. What is the mean and standard deviation of the $ave_z$ variable?

    a. mean = 0; standard deviation = 1
    b. mean = 0; standard deviation approximately 1
    c. mean = 0; standard deviation > 1
    d. Cannot be determined from the information given

```{r}
Answer: c
```

```{r}
x <- c(0, -1, 1, -5, 10)
y <- c(-2, -3, 5, -7, 7)
z_x <- scale(x)
z_y <- scale(y)
my_z <- data.frame(z_x, z_y)
library(dplyr)
options(scipen=999)
my_z %>% 
  mutate(ave_z = z_x + z_y / 2) %>% 
  summarise_each(funs(mean, sd))
```

24. The sampling distribution and standard error of a statistic can be calculated by:

    a. exhaustive and exact calculations where formula solutions are possible
    b. simulation
    c. formula approximations
    d. repeatedly taking random samples of a given size from a population
    e. all of the above

```{r}
Answer: e
```

---

For questions 25-32 suppose you have four children in a reading group (Beth, Marianne, Steven, Joel) and you randomly pick one child to lead the discussion in group each day of a 5-day week. Furthermore, we define the outcome of each day's selection to be binary: Steven leads the discussion or he doesn't.

25. The number of times Steven leads the discussion in a week would be the:

    a. probability distribution for this experiment
    b. probability of an outcome
    c. constant in this experiment
    d. random variable in this experiment
    
```{r}
Answer: d
```

26. The probability that Steven leads the discussion all 5 days in a week is the:

    a. expected value
    b. probability distribution
    c. probability of a simple event
    d. random variable
    
```{r}
Answer: c
```

27. If we could replicate this experiment many, many times, the average number of times that Steven leads the discussion in a week would be the:

    a. expected value
    b. probability distribution
    c. probability of an outcome
    d. random variable
    
```{r}
Answer: a
```

28. If we found the probability of Steven leading the discussion zero times, one time, two times, three times, four times, and five times, the set of six probabilities would be the:

    a. expected value
    b. probability distribution
    c. probability of an outcome
    d. random variable
    
```{r}
Answer: b
```

29. \* Monday's selection of a discussion leader could be considered a:

    a. binomial experiment
    b. bernoulli trial
    c. neither
    
```{r}
Answer: b
```

30. \* The selections of discussion leaders for the week constitute a:

    a. binomial experiment
    b. bernoulli trial
    c. neither
    
```{r}
Answer: a
```

31. What is the probability that Steven would be selected all 5 days of the week?

    a. .00000000000000
    b. .0009765625
    c. .0039065500
    d. .25
    
```{r}
Answer: b
```

```{r}
.25^5
dbinom(5, 5, .25) #or
1 - pbinom(4, 5, .25) #or
pbinom(4, 5, .25, lower.tail = FALSE)
```

32. What is the expected number of times that Steven would be selected?

    a. 0.00
    b. 1.00
    c. 1.25
    d. 2.50
    
```{r}
Answer: c
```

```{r}
n <- 5
p <- .25
n*p # definition of expected value for a binomial rv
```
---





